Fixing Vulnerabilities Hinders Maintainability

### 2019-02-04 paper/ICPC19/icpc19.tex
S1 and S2

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines 
or formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which suffered
a lot of criticism regarding the costs associated and poor technical evaluation; 
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is 
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security which we believe it was one of the
first steps taken by SIG to introduce security on their maintainability model.

@@

The particular subject of this paper is to explore whether there is a trade-off
between refactoring source code to patch disclosed vulnerabilities while keeping
(or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
groups of refactored elements and non-refactored and conclude that the source
code elements subjected to refactorings had significantly lower maintainability
than the elements not affected.

@@

Every SSL/TLS connection begins with a handshake which is responsible by the
negotiation between two parties. DTLS provides a mechanism for fragmenting a
handshake message over a number of packets, each of which can be transmitted
separately. In the example (Listing~\ref{lst:vuln}), all functions receive as
input the SSL session, \texttt{s}. In the \texttt{dtls1\_reassemble\_fragment}
function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is
checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
$13$) and then against the maximum handshake message size allowed by DTLS1 for
\texttt{s} (line $16$-$21$). However, it is not verified if
\texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line
$37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
than the allowed might be allocated. According to the message commit, $10$
handshake messages are allowed per DTLS connection. Hence, a remote attacker
could leverage this issue to trigger memory allocations corresponding to large
length values and lead the system to memory exhaustion that could end in a DoS
attack. The code changes performed to fix the vulnerability were the following:

~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
inserted to return the number of maximum number of bytes allowed in a DTLS
handshake message for \texttt{s}. The minimum is $16$KB but may be greater if
the maximum certificate list size requires it which is set by the user. More
specifically, it checks if the maximum certificate list size
(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
(\texttt{max\_len}). If that condition holds, the function returns the size of
the certificate list, otherwise, it returns the maximum length. This function
replaces lines $18$-$21$.

~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to
ensure that the message and the fragment length are never higher than the
maximum allowed by DTSL. However, \ref{lst:func5} is responsible for patching
the vulnerability.

Although checking the maximum size of a handshake message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the codebase which produced a negative impact in the
project maintainability. Even though the developer tries to simplify the
codebase with the creation of a smaller unit
\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it also disrupts
$2$ of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which
contributes to a potential violation of the \emph{Keep Your Codebase Small}
guideline. Hence, we have observed that the impact of this refactoring on the
OpenSSL maintainability was negative.

@@

There are security flaws that are more difficult to refactor than others. For
instance, implementing secure authentication is not as easy as fixing a
cross-site scripting vulnerability since usually the last one is can be fixed
without adding new lines of code, as seen in Figure~\ref{lst:fix}. The developer
added the function \emph{htmlentities} to escape the data inserted by the
variable \emph{\$\_['file']}. Knowing which patterns are more likely to increase
maintainability issues is one step forward to bring awareness to security
engineers of what patterns need more attention. The taxonomy of security
patterns used to answer this question is the same as the one provided by the
authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
separately for each pattern.

>>>>

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines
or formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model).

@@

The particular subject of this paper is to explore whether there is a trade-off
between refactoring source code to patch disclosed vulnerabilities and keeping
(or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
groups of refactored elements and non-refactored and conclude that the source
code elements subjected to refactorings had significantly lower maintainability
than the elements not modified.

@@

Every SSL/TLS connection begins with a handshake which is responsible by the
negotiation between two parties. DTLS provides a mechanism for fragmenting a
handshake message over a number of packets, each of which can be transmitted
separately. In the example (Listing~\ref{lst:vuln}), all functions receive as
input the SSL session, \texttt{s}. In the \texttt{dtls1\_reassemble\_fragment}
function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is
checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
$13$) and then against the maximum handshake message size allowed by DTLS1 for
\texttt{s} (line $16$-$21$). However, it is not verified if
\texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line
$37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
than the allowed might be allocated. According to the commit message, $10$
handshake messages are allowed per DTLS connection. Hence, a remote attacker
could leverage this issue to trigger memory allocations corresponding to large
length values and lead the system to memory exhaustion that could end in a DoS
attack. The code changes performed to fix the vulnerability were the following:

~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
inserted to return the maximum number of bytes allowed in a DTLS
handshake message for \texttt{s}. The minimum is $16$KB but may be greater if
the maximum certificate list size requires it which is set by the user. More
specifically, it checks if the maximum certificate list size
(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
(\texttt{max\_len}). If that condition holds, the function returns the size of
the certificate list, otherwise, it returns the maximum length. This function
replaces lines $18$-$21$.

~\ref{lst:func5} and~\ref{lst:func4} are guards added to different functions to
ascertain that the message and the fragment length are never higher than the
maximum allowed by DTSL. In particular, \ref{lst:func5} is responsible for patching
the vulnerability.

Although checking the maximum size of a handshake message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the codebase which produced a negative impact in the
maintainability. Even though the developer tries to simplify the
codebase by introducing a smaller unit
\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change also
disrupts $2$ of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which
contributes to a potential violation of the \emph{Keep Your Codebase Small}
guideline. Hence, we have observed that the impact of this refactoring on the
OpenSSL maintainability was negative.

@@

There are security flaws that are more difficult to refactor than others. For
instance, implementing secure authentication is not as easy as fixing a
cross-site scripting vulnerability since usually the last one is can be fixed
without adding new lines of code, as seen in Figure~\ref{lst:fix}. The developer
added the function \emph{htmlentities} to \textit{escape} the data given by the
variable \emph{\$\_['file']}. Knowing which patterns are more likely to increase
maintainability issues is one step forward to bring awareness to security
engineers of what patterns need more attention. The taxonomy of security
patterns used to answer this question is the same as the one provided by the
authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
separately for each pattern.
